Advances in Artificial Intelligence for Data Visualization: Developing Computer Vision Models to Automate Reading of Data Plots, with Application to Predictive Model Diagnostics

### 2024-08-26 Thesis/04-chap4.qmd
make shiny -> Shiny, and add right citation for ecostats

Regression software is widely available today, but tools for effective diagnostics are still lagging. Perhaps, one of the reasons is that it is hard. Conventional tests, that would make the task easy, are often too sensitive, which would result in adequate models being abandoned if the analyst strictly adhered to the test decision. The recommended advice is to have the analyst assess the strength of patterns in residual plots. This requires human effort and also suffers from the potential for inconsistent decisions from different analysts. Using a lineup protocol can help to alleviate inconsistency, but requires even more human effort. This is the type of task where a robot might be employed to do the disagreeable work that currently requires a human. Here we describe a new R package that includes a computer vision model for automated assessment of residual plots, and an accompanying shiny app for ease of use. For a user-provided sample of residuals, it predicts a measure of visual signal strength (VSS) and provides a suite of supporting information to assist the analyst diagnose their model fit.

@@

Enhanced visual diagnostics forms the second group, which offer advanced visual aids for interpreting diagnostic plots. This includes `ecostats` [@ecostats], which adds simulation envelopes to  residual plots, and `DHARMa` [@dharma], which compares the empirical 0.25, 0.5, and 0.75 quantiles in scaled residuals with their theoretical counterparts. `DHARMa`'s focus is detecting model violations such as heteroscedasticity and incorrect functional forms, and uncovering issues specific to generalized linear models and mixed-effect models, like over/under-dispersion. <!--By offering these enhanced visualizations, `DHARMa` enables users to more easily identify potential issues in their regression models that might not be immediately apparent with standard diagnostic plots. In summary, this group of packages enhances the interpretability of diagnostic plots by drawing attention to critical elements such as trends, clusters, and outliers.--> It also adds results from conventional tests as annotations, labels, or text within the plot, to assist in avoiding misinterpretation.

>>>>

Regression software is widely available today, but tools for effective diagnostics are still lagging. Perhaps, one of the reasons is that it is hard. Conventional tests, that would make the task easy, are often too sensitive, which would result in adequate models being abandoned if the analyst strictly adhered to the test decision. The recommended advice is to have the analyst assess the strength of patterns in residual plots. This requires human effort and also suffers from the potential for inconsistent decisions from different analysts. Using a lineup protocol can help to alleviate inconsistency, but requires even more human effort. This is the type of task where a robot might be employed to do the disagreeable work that currently requires a human. Here we describe a new R package that includes a computer vision model for automated assessment of residual plots, and an accompanying Shiny app for ease of use. For a user-provided sample of residuals, it predicts a measure of visual signal strength (VSS) and provides a suite of supporting information to assist the analyst diagnose their model fit.

@@

Enhanced visual diagnostics forms the second group, which offer advanced visual aids for interpreting diagnostic plots. This includes `ecostats` [@warton_global_2023], which adds simulation envelopes to  residual plots, and `DHARMa` [@dharma], which compares the empirical 0.25, 0.5, and 0.75 quantiles in scaled residuals with their theoretical counterparts. `DHARMa`'s focus is detecting model violations such as heteroscedasticity and incorrect functional forms, and uncovering issues specific to generalized linear models and mixed-effect models, like over/under-dispersion. <!--By offering these enhanced visualizations, `DHARMa` enables users to more easily identify potential issues in their regression models that might not be immediately apparent with standard diagnostic plots. In summary, this group of packages enhances the interpretability of diagnostic plots by drawing attention to critical elements such as trends, clusters, and outliers.--> It also adds results from conventional tests as annotations, labels, or text within the plot, to assist in avoiding misinterpretation.
